`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages

### 2016-04-28 response-to-reviewers.tex

Here you can describe the analysis steps that are common to both datasets. Please provide brief descriptions of models M1 and M8. I note that in Lewis' Mk model, the k stands for the dimension of the character substitution matrix, so M2 means a 2x2 substitution matrix (binary characters). This does not seem to be the case here. The authors could either change their notation slightly (suggestion: Mk-1, Mk-8), or explicitly state that their nomenclature does not follow Lewis.

\response{We added a section called Phylogenetic analyses with following text:}

\respquote{Following \cite{gavryushkina2015bayesian}, we apply the Lewis Mk model \cite{Lewis2001} for discrete morphological character evolution, which assumes a character can take $k$ states and the transition rates from one state to another are equal for all states. As did \cite{gavryushkina2015bayesian}, we treat 34 characters that were ordered in \cite{ksepka2012} as unordered. 
We applied two phylogenetic models to the penguin data set: Mk-1 and Mk-8. Mk-1 assumed a strict morphological clock and no gamma-distributed rate heterogeneity among sites. Model Mk-8 partitioned the alignment into six partitions, with the $i$'th partition containing all characters that had $k=i+1$ character states across the sampled taxa.}

Provide a better description of the stepwise analysis of fossil ages.

\response{We have clarified details of the stepwise analysis of the fossils in the new ``Phylogenetic analyses'' section.}

My impression (but I am not totally sure) is that in one analysis, fossil ages are unknown parameters and have uniform prior distributions set according to the stratigraphic bounds.

\response{No, they have prior distributions set according to the fossilized birth-death prior, conditioned on the stratigraphic bounds. This is detailed in the new ``Phylogenetic analyses'' section. }

Then, in stepwise analysis, each fossil age is in turn represented not by the stratigraphic uniform distribution, but by a prior density derived from the fossilised birth-death process.

\response{No, each fossil is in turn analyzed using the prior density derived from the fossilised birth-death process without conditioning on its stratigraphic bounds, while the remaining fossils are conditioned on their stratigraphic bounds. This is detailed in the new ``Phylogenetic analyses'' section. }

 Give some details please.

 \response{We have added the opening three paragraphs in the new ``Phylogenetic analyses'' section that give the details.}

  I was trying to work out the general form of the posterior distribution, I was guessing something like:

@@

$$p(tip\_ages) p(node\_ages) L(D|ages,tree)$$

where $p(tip\_ages)$ is the prior density on fossil ages using uniform distributions, $p(node\_ages)$ is the prior on internal node ages given the BD process, and $L(D|ages,tree)$ is the likelihood of the morphological data D, given the ages and the topology calculated using Lewis (2001) formula. However, it seems from the text in p.5 that another likelihood term may be involved. For the stepwise deletion procedure, I am guessing the priors are something like $p(tip\_ages\hat{}) p(node\_ages, tip\_age*)$, where now $tip\_ages\hat{}$ includes all the fossil ages except the one under analysis, which has been moved into the other BD-based prior term $(tip\_age*)$. Please provide general forms for the posteriors and make the connection with equation (1).

\response{We have now provided a general form of the posterior we are sampling under the new section on ``Phylogenetic analyses''. }

@@

Give some more info on how the prior on the root age is set. Is Tmax the same as the root age? If it is, please state explicitly.

\response{$T$ is not the age of the root, it is the origin of the fossilized birth-death process, and is older than the root.}

@@

\response{We have added the prior information for the morphological substitution rate and log standard deviation of the relaxed clock.}

p.8 paragraph starting on line 43. Do you mean that with the particular implementation of marginal likelihood calculation in BEAST you cannot calculate the BF? Or is this a general problem with any implementation of stepping-stones and this model? I am also confused, do sampling times refer to the ages of fossil tips? Aren't these unknown parameters being estimated in the MCMC? Then they would appear as terms in the prior and the likelihood. (?)

\response{The ages of the fossil tips are indeed unknown parameters estimated by the MCMC. Their prior distribution emerges from the marginal distribution of the fossilized birth death prior of tip-timed trees, which can be sampled using MCMC, but which we don't have an analytical solution for.}

p.9 Results section. For both data sets the authors state that the relaxed clock model fits the data better, but I can't find the BF for this.

\response{We didn't provide BF. For the penguin data set the BF was provided in \cite{gavryushkina2015bayesian}.}

Also, for both data sets, the authors have sections titled "Penguins conform well to a strict morphological clcok" and "Canids conform well to a morphological clock", but these headings seem to contradict the statement in lines 48-49 p.12 of discussion. I am also puzzled as the authors have not quantified here departure from the clock.

\response{We mean that the data sets conform to a relaxed morphological clock, not a strict morphological clock. A relaxed morphological clock is still a morphological clock, just with greater variance.}

@@

Please provide a table with prior and posterior mean estimates of the age of the root for both data sets under all models used, and 95\% CI. Similarly, in the same table provide prior and posterior estimates of mean morphological substitution rate for all data sets and models (relaxed and strict clock). For the relaxed-clocks, provide prior and posterior estimates of rate-variation parameter. Then, when the authors discuss whether the data fit or not the strict clock, they should refer to the estimates in the table.

\response{
\begin{table}[ht]
\centering
\begin{tabular}{rrrr}
  \hline
Parameter & Penguins Mk-1 & Penguins Mk-8 & Canids Mk-1 \\ 
  \hline
$t_{MRCA}$ & & & \\
morph. clock rate & & & \\
log std. dev. rate & & & \\
\hline
  \end{tabular}
\caption{Summary of key parameters for the three main analyses.} 
\label{fossilTable1}
\end{table}
}

### 2016-04-28 fossilDating-revision.tex
morphological clocks

In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model. 
We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock. 
We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 
This may suggest that the larger reference set of fossils in the canid data set (115 versus 35) more than makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates. However since the average age of the canid fossils is considerably less than that for the penguin fossils, it could simply be that the absolute errors are smaller because the true values are smaller.

>>>>

In this paper we have demonstrated that even a small number of morphological characters (some of the fossils had as few as 7 morphological traits coded) can be used in the context of a rich fossil reference data set, to provide an accurate age of the fossil based on a phylogenetic model.

\subsection*{Morphological clock}

We found that although a strict morphological clock does a surprisingly good job of estimating fossil ages, there is evidence that phylogenetic estimation of fossil ages is improved by a model that includes a relaxed morphological clock.

%AJD revision
There is a long history of the study of the evolutionary rates of phenotypic characters \cite{Simpson1949, Gingerich1981}. Phenotypic evolution is typically characterized as being less clock-like than molecular evolution \cite{Kimura}. While there are many examples of extremely slow and fast rates of phenotypic evolution in the literature, this is also true for molecular rates. We are not aware of a systematic comparison of variation in rates at the phenotypic and molecular levels. Regardless, variation in the rate of phenotypic evolution across the phylogeny can be accommodated with a relaxed morphological clock, as we used in model \Mrelaxed{}. In the penguin data set the log standard deviation of the relaxed morphological clock was comparable to values obtained for molecular clocks.   
%%%

%http://earth.geology.yale.edu/~ajs/1993/11.1993.17Gingerich.pdf

We found that fossil estimates were both slightly more accurate and more precise on average in the canid data set. 
This may suggest that the larger reference set of fossils in the canid data set (115 versus 35) more than makes up for the smaller number of characters (122 versus 245) with regards to accuracy and precision of fossil age estimates. However since the average age of the canid fossils is considerably less than that for the penguin fossils, it could simply be that the absolute errors are smaller because the true values are smaller.
